\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2020-06-29 IPANEMAP-2019.tex
Updates from Overleaf

\noindent We introduce \OurTool{}, an automated method for the modeling of RNA structure from multiple probing reactivity profiles. Input profiles can result from experiments based on diverse protocols, reagents, \Rev{or collection of variants}, and are jointly analyzed to predict the dominant conformations of an RNA.

\noindent \OurTool{} combines sampling, \Rev{clustering}, and multi-optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. The analysis of multiple reactivity profiles, both publicly available and produced in our study, demonstrates the good performances of \OurTool, even in a mono probing setting. It confirms the potential of integrating multiple sources of probing data, informing the design of informative probing assays.

@@

Reagents used within \Rev{selective 2' -hydroxyl acylation analyzed by primer extension (SHAPE)~\cite{Wilkinson2006} protocols} represent a popular class of probes. They react with the 2’-hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides involved in stable interactions. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamic tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates Adenines and Cytosines if not involved in a hydrogen bond, while   CMCT reacts with the Watson-Crick face of unpaired Guanosines and Uracyls~\cite{Ehresmann1987,Brunel2000}. These two reagents not only reveal Watson-Crick base pairing, but also other types of contacts involving the same edge. The diversity of probes\Rev{, some of which are usable \emph{in vivo},} not only increases coverage of the different positions and structural contexts, \Rev{but also provides different qualitative information. \Rev{Modeling can also benefit from a joint analysis of reactivity profiles of single-point mutants, assuming structural homology~\cite{Cordero2012a}}. Integration of multiple sources of probing to improve structure prediction  has thus been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017} but, to the best of our knowledge, has never been fully automated.}

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy \Rev{to assess the unpaired nature of individual nucleotides}. The past couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the used reagent/enzyme, significantly unreactive or reactive positions were forced to remain paired or  unpaired within predicted models. However, such \emph{hard constraints} can be overly  sensitive to the choice of the cutoff, leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing~\cite{Smola2015}, new methods~\cite{Deigan2009,Zarringhalam2012,Washietl2012} lifted the requirement of a threshold, supplementing  Turner nearest-neighbor energy model~\cite{Turner2010} with pseudo-energies derived from the reactivities. They  then performed a (pseudo) energy minimization, optimizing a tradeoff between the thermodynamic stability and its compatibility with the reactivity profile. Popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. \Rev{Besides free-energy optimization, such methods notably} include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signals produced by various \Rev{probing experiments to predict one or several secondary structure models for a given RNA. It takes as input one or several reactivity profiles produced in various \Def{conditions}, broadly defined to denote the conjunction of a reagent, a probing technology, ionic concentrations and, in extreme cases, structurally homologous mutants.}

@@

	\caption{\OurTool{} workflow: \OurTool{} takes as input an RNA sequence with profiling data, denoted by reactivities, from various experimental conditions. \OurTool{} proceeds, first,  with a stochastic sampling that results into samples of predicted secondary structure. The data-driven predicted structures are then gathered in one sample, serving as input for the clustering step. \OurTool{} proceeds, then, with an iterative clustering that ends once \Rev{a stopping criterion is} reached. This step allows to identify the adequate number of clusters $k$ to be considered. The $k$ resulting clusters are then \Rev{represented} by their centroid structures. Clusters figuring on the 2D-Pareto frontier are considered to be optimal and subsequently their corresponding centroids are reported as the predicted structure through \OurTool.}\label{fig:approach}
\end{figure}

\paragraph{Sampling the pseudo-Boltzmann ensembles.} Our method, summarized in Figure~\ref{fig:approach}, takes as input a set $\mathcal{D}$ of probing experiments, each materialized by a reactivity profile. It starts by producing (multi)sets of representative structures for each of the reactivity profiles using a SHAPE-directed variant of the classic Ding-Lawrence algorithm~\citep{Ding2003}. Following Deigan\etal\cite{Deigan2009}, \Def{soft constraints} are used to complement the free energy contributions of the classic Turner energy model with pseudo energy contributions resulting from the reactivity derived from a probing experiment.  Given a reactivity $r_i$ for a position $i$ in a probing $d$, we associate a free-energy bonus to unpaired positions, defined as
$$\Delta G_d(i) = m \log(r_i +1 )+b$$ 
using $m=1.3$ and  $b=-0.4$. \Rev{Those values were halved in comparison to those recommended by Deigan\etal\cite{Deigan2009}, following a grid search optimization on the Cordero dataset, and based on the rationale that lower absolute values for pseudo-energy bonuses increase the expected overlap between pseudo-Boltzmann ensembles.} Those pseudo energy contributions effectively guide our predictions towards a subset of structures that are in good agreement with probing data. For each condition $d$ in $\mathcal{D}$, we use the soft constraints framework~\cite{Lorenz2016} of the \Software{RNAfold} software~\cite{Lorenz2011} to produce a random (multi)set $\mathcal{S}_d$ of $\NumStruct$ structures in the pseudo-Boltzmann ensemble.

\Rev{\paragraph{Clustering across conditions.} In order to infer recurrent conformations across sampled sets, \OurTool{} agglomerate structure (multi)sets while keeping track of their condition of origin, and \Def{clusters} with respect to the \Def{base-pair distance}, the number of base pairs differing between two structures. A clustering algorithm then partitions the (multi)set of sampled structures into \emph{clusters}, (multi)sets of structures such that the accumulated sum of distances over clusters is minimized.

@@

Moreover, define the (pseudo-)\Def{Boltzmann condition probability} of a structure $S$, generated for a probing condition $d$ as part of a sampled  set $\mathcal{S}_d$, as
$$\mathbb{P}_d(S) = \frac{e^{-E_d(S)/RT}}{\mathcal{Z}^*_d} \text{, with } \mathcal{Z}^*_d := \sum_{S'\in \mathcal{S}_d} e^{-E_d(S')/RT}$$
where $E_d(S)$ is the pseudo free-energy, including the Turner free-energy, assigned to the structure $S$ within the probing condition $d$.
The \Def{stability} of a cluster $C$ denoted its accumulated pseudo-Boltzmann probability across conditions, computed as
$$\text{Stability}(C) = \sum_{d\in \mathcal{D}} \sum_{S\in C\cap \mathcal{S}_d}  \mathbb{P}_d(S).$$
A cluster is deemed \Def{significantly populated} if its stability exceeds a predefined threshold $\epsilon$. 
We set  $\epsilon =\Rev{|\mathcal{D}|}/3$ by default, such that at most three clusters are deemed significantly populated, and used as our primary candidates.
Finally, we consider two clusters to be \Def{highly similar} if their centroid structures differ by at most $\delta$ base pairs ($\delta=1$ by default), allowing the identification of clusters in the presence of minor variations.

@@

\paragraph{Filtering the promising conformer(s).} Finally, we \Def{elect the most promising cluster(s)}, and return their centroid(s). While the final number of clusters may potentially be large, only a handful of clusters are expected to represent structures that \Rev{are} both stable, and supported by a large number of conditions. The remaining clusters are indeed probably artifacts of the clustering method, but nevertheless useful to filter out \emph{noisy} structures.

\Rev{We postulate that a perfect cluster should have large stability, as defined above, and be representative of several conditions.} In the presence of a set of experimental conditions $\mathcal{D}$, we consider that a cluster $C$ supports a given condition $d$ when its probability within $d$ exceeds a given threshold $\tau$. The number of conditions supporting a cluster $C$ is defined as
$$\text{Support}(C) = \left|\left\{d\in \mathcal{D} \mid \left(\textstyle\sum_{S\in C\cap \mathcal{S}_d} \mathbb{P}_d(S)\right)\ge \tau \right\}\right|.$$
The value of $\tau$ is set to $1/(\NumClust+1)$ \Rev{with $\NumClust$ the final number of clusters},  ensuring at least one supporting cluster for each condition.

@@

\paragraph{Hajdin dataset.} 
A dataset was gathered by Hajdin\etal\cite{Hajdin2013} to validate the predictive capacities of probing data-driven predictions. It consists of $24$ RNA sequences with known secondary structures for which a single chemical probing reactivity profile (1M7-\SH) was made available.
This dataset includes sequences originating from a variety of organisms, and spans lengths ranging from $34$ nts to $530$ nts, with a focus on riboswitches and complex RNA architectures \Rev{(full list in Supp. Table S3)}.

@@

\paragraph{Didymium structural model and probing data (\didy{} dataset).} 
We considered the 188 nucleotides Lariat capping ribozyme from {\itshape Didymium iridis}, resolved 3.85 \AA{} resolution using X-ray cristallography (PDB: 4P8Z)~\citep{Meyer2014}.  We annotated the secondary structure elements using the \Software{DSSR} software from the 3DNA suite~\cite{Lu2015}. \Rev{Non-canonical base pairs were removed, and a non-pseudoknotted secondary structure was extracted as the maximum subset of non-crossing base pairs~\citep{Smit2008}.}

@@

\Rev{To systematically assess the potential of multiple sources of probing, we considered a difficult example, the \textit{Didymium iridis} Lariat Capping ribozyme (\didy{}). The native structure of \didy{}, shown in Figure~\ref{fig:didy} is highly complex, and features two pseudoknots which cannot be explicitly modeled by most computational methods, making \didy{} a challenging target for secondary structure prediction.}

\didy{} was probed with different SHAPE reagents: 1M7 (1-methyl-7 nitrosatoic anhydre), NMIA (N-Methyl Isatoic Anhydre), BzCN (Benzoyl cyanide) and NAI (2-methylnicotinic acid imidazolide) in presence and absence of Mg$^\text{2+}$. \Rev{\didy{} was also probed with DMS (DiMethylSulfate) and CMCT, in presence of Mg$^\text{2+}$, resulting in a total of 16 \Def{probing conditions}, a term we use in the following to denote a combination of probing technology, reagent and presence/absence of Mg$^\text{2+}$ (+ sequencing technology). For each probing condition, three experiments were performed in presence/absence of the reagent, and in a denatured context, following classic SHAPE protocols~\cite{Wilkinson2006,Smola2015}.}

@@

The Matthews Correlation Coefficient (MCC) is a classic metric for assessing the quality of a predicted structure $S$, identified by a set of base pairing positions, compared to an accepted reference (native) structure $\Ref$. It represents a compromise \Rev{between the main metrics derived from the confusion matrix}, and is defined as 
$$\text{MCC}(S\mid \Ref)=\frac{\text{TP} \times \text{TN} -\text{FP} \times \text{FN}}{\sqrt[]{(\text{TP}+\text{FP})(\text{TP}+\text{FN})(\text{TN}+\text{FP})(\text{TN}+\text{FN})}},$$
where TP, FP, TN, FN represent the correctly/erroneously predicted and correctly/erroneously omitted base pairs in $S$ with respect to $\Ref$.
\Rev{MCC is a stringent metric, taking values between -100\% and 100\%, 0\% being the expected MCC of a "coin tossing" random predictor.}

@@

\Rev{
\subsection*{Assessment of statistical significance}
Following Xu\etal\cite{Xu2012}, we assessed the statistical significance of observed changes in predictive performances using a two-tailed Student $t$-test (unequal variance) with usual significance threshold $\alpha=5\%$. When comparing performances within a given dataset, we used the paired version of the test.

@@

\subsection*{Considering multiple probing conditions \Rev{appears to} improve the quality of predictions}\label{sec:cordero}

@@

	\caption{Dot-bracket representations of the native structure and \OurTool{} predictions for the glycine riboswitch sequence of the Cordero\etal\cite{Cordero2012} data set, supplemented with an increasing collection of probing conditions. \Rev{Please note that, throughout the manuscript, by "conditions" we mean a combination of the nature of the probe, the probing technology, the ions present, the temperature, or any other "environmental" variation} Positions with green and red backgrounds indicate correctly and incorrectly predicted base pairs.}\label{fig:glycine_example}

@@

	\caption{\Rev{Predictive performances (GM) of \OurTool{} on the Cordero~\cite{Cordero2012a} dataset consisting of reactivities associated with the CMCT, NMIA and DMS probing for 6 reference RNAs. (Pseudo) MFE and MEA structures, predicted using \Software{RNAfold}~\cite{Lorenz2011}, are included as references in the case of a single condition, and in the absence of probing.}}
	\label{cordero1}
\end{figure}

@@

Remarkably, the combination of the three conditions leads to the best overall predictions, averaging $78.5\%$ GM.
This \Rev{seemingly} improves by $8.5\%$, $2.7\%$ and $1.4\%$ over the average predictions achieved using CMCT, DMS and NMIA respectively.
%CMCT/DMS/NMIA :0.6988333333/0.7588333333/0.771
Table~\ref{fig:glycine_example} illustrates the incremental refinement of \OurTool predictions for a glycine riboswitch upon increasing the number of considered probing conditions.
\Rev{However, these observations merely suggest an improvement, as statistical significance cannot be established on such a small dataset.}
Still, they suggests some level of complementarity between probing conditions, as already suggested by recent analyses~\citep{Brunel2000,Steen2012,Rice2014,Yu2018}  and supported by further analyses in this paper.

@@

\subsection*{\OurTool outperforms state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assessed the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compared its performances against \Software{RSample} software recently introduced by Spacic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform \Rev{comparably as} a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAStructure} suite~\citep{Reuter2010}.
We considered the Hajdin dataset~\cite{Hajdin2013}, which consists of 24 RNAs of lengths ranging from 47 to 500 nts, all believed to fold into a unique documented conformation.

\begin{figure}
	\includegraphics[width=\linewidth]{graphs/RsampleVsIPANEMAP/Accuracy}
	\caption{\Rev{Comparison of predictive performances (GM) achieved by  \OurTool{} and \Software{Rsample} over the Hajdin\etal\cite{Hajdin2013} data set, consisting of 24 RNAs which associated SHAPE reactivities. The length of each individual RNA is indicated within square brackets.} \label{fig:Vsrsample}}
\end{figure}

\Rev{We used the default number of sampled structures for \Software{RSample}, namely 10~000 to compute correction factors and 1~000 for the clustering phase, and \OurTool (1~000 structures).  For all RNAs, a single structure was returned by both software. Note that the Pareto-optimality always implies a single returned cluster/structure for \OurTool{} in a mono probing setting.}
We computed and report in Figure~\ref{fig:Vsrsample} the individual Geometric Means (GM) of predicted structures. \Rev{Sensitivity and PPV can be found in Supp. Table S4, and Supp. Table S5 reports the stability of predictions over 10 independent runs of \OurTool.}

%\paragraph{Results description}
A first observation is that \OurTool generally outperforms \Software{RSample}, averaging $80\%$ GM as opposed to $68\%$ for its competitor. \Rev{The superiority on \OurTool on this benchmark appears statistically significant according to a two-tailed paired $t$-test (P-value$=6\times10^{-3}$) with observed data.}
Out of the 24 considered RNAs, \OurTool outperforms its competitor for 16 out of the 24 RNAs, achieves similar performances for 2 of them, and is dominated for the remaining 6 cases. Whenever dominant, \OurTool achieves an average GM score of $82\%$, dominating by $32.8\%$ its competitor, while being dominated by $3.5\%$ over the remaining examples (including ties). 
%Difference<-c(IPANEMAP-Rsample)
%> sum(Difference[1:18])
%[1] 3.17
%> sum(Difference[19:24])
%[1] -0.382
Remarkably, \OurTool predicts secondary structures much more consistently than its competitor. 
It ensures GM that are greater than 50\% in all cases, and even exceeds 60\% over all except for the yeast tRNA, encoding the Aspartic acid and HIV1 5' pseudoknot domain. The standard deviation of GM for \OurTool is $12.4\%$, compared to the larger value of $21\%$ for \Software{RSample} predictions.

>>>>

\noindent We introduce \OurTool{}, an automated method for the modeling of RNA structure from multiple probing reactivity profiles. Input profiles can result from experiments based on diverse protocols, reagents, \RevA{or collection of variants}, and are jointly analyzed to predict the dominant conformations of an RNA.

\noindent \OurTool{} combines sampling, \RevA{clustering}, and multi-optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. The analysis of multiple reactivity profiles, both publicly available and produced in our study, demonstrates the good performances of \OurTool, even in a mono probing setting. It confirms the potential of integrating multiple sources of probing data, informing the design of informative probing assays.

@@

Reagents used within \RevA{selective 2' -hydroxyl acylation analyzed by primer extension (SHAPE)~\cite{Wilkinson2006} protocols} represent a popular class of probes. They react with the 2’-hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides involved in stable interactions. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamic tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates Adenines and Cytosines if not involved in a hydrogen bond, while   CMCT reacts with the Watson-Crick face of unpaired Guanosines and Uracils~\cite{Ehresmann1987,Brunel2000}. These two reagents not only reveal Watson-Crick base pairing, but also other types of contacts involving the same edge. The diversity of probes\RevA{, some of which are usable \emph{in vivo},} not only increases coverage of the different positions and structural contexts, \RevA{but also provides different qualitative information. \RevA{Modeling can also benefit from a joint analysis of reactivity profiles of single-point mutants, assuming structural homology~\cite{Cordero2012a}}. Integration of multiple sources of probing to improve structure prediction  has thus been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017} but, to the best of our knowledge, has never been fully automated.}

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy \RevA{to assess the unpaired nature of individual nucleotides}. The past couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the used reagent/enzyme, significantly unreactive or reactive positions were forced to remain paired or  unpaired within predicted models. However, such \emph{hard constraints} can be overly  sensitive to the choice of the cutoff, leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing~\cite{Smola2015}, new methods~\cite{Deigan2009,Zarringhalam2012,Washietl2012} lifted the requirement of a threshold, supplementing  Turner nearest-neighbor energy model~\cite{Turner2010} with pseudo-energies derived from the reactivities. They  then performed a (pseudo) energy minimization, optimizing a tradeoff between the thermodynamic stability and its compatibility with the reactivity profile. Popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. \RevA{Besides free-energy optimization, such methods notably} include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signals produced by various \RevA{probing experiments to predict one or several secondary structure models for a given RNA. It takes as input one or several reactivity profiles produced in various \Def{conditions}, broadly defined to denote the conjunction of a reagent, a probing technology, ionic concentrations and, in extreme cases, structurally homologous mutants.}

@@

	\caption{\OurTool{} workflow: \OurTool{} takes as input an RNA sequence with profiling data, denoted by reactivities, from various experimental conditions. \OurTool{} proceeds, first,  with a stochastic sampling that results into samples of predicted secondary structure. The data-driven predicted structures are then gathered in one sample, serving as input for the clustering step. \OurTool{} proceeds, then, with an iterative clustering that ends once \RevA{a stopping criterion is} reached. This step allows to identify the adequate number of clusters $k$ to be considered. The $k$ resulting clusters are then \RevA{represented} by their centroid structures. Clusters figuring on the 2D-Pareto frontier are considered to be optimal and subsequently their corresponding centroids are reported as the predicted structure through \OurTool.}\label{fig:approach}
\end{figure}

\paragraph{Sampling the pseudo-Boltzmann ensembles.} Our method, summarized in Figure~\ref{fig:approach}, takes as input a set $\mathcal{D}$ of probing experiments, each materialized by a reactivity profile. It starts by producing (multi)sets of representative structures for each of the reactivity profiles using a SHAPE-directed variant of the classic Ding-Lawrence algorithm~\citep{Ding2003}. Following Deigan\etal\cite{Deigan2009}, \Def{soft constraints} are used to complement the free energy contributions of the classic Turner energy model with pseudo energy contributions resulting from the reactivity derived from a probing experiment.  Given a reactivity $r_i$ for a position $i$ in a probing $d$, we associate a free-energy bonus to unpaired positions, defined as
$$\Delta G_d(i) = m \log(r_i +1 )+b$$ 
using $m=1.3$ and  $b=-0.4$. \RevA{Those values were halved in comparison to those recommended by Deigan\etal\cite{Deigan2009}, following a grid search optimization on the Cordero dataset, and based on the rationale that lower absolute values for pseudo-energy bonuses increase the expected overlap between pseudo-Boltzmann ensembles.} Those pseudo energy contributions effectively guide our predictions towards a subset of structures that are in good agreement with probing data. For each condition $d$ in $\mathcal{D}$, we use the soft constraints framework~\cite{Lorenz2016} of the \Software{RNAfold} software~\cite{Lorenz2011} to produce a random (multi)set $\mathcal{S}_d$ of $\NumStruct$ structures in the pseudo-Boltzmann ensemble.

\RevA{\paragraph{Clustering across conditions.} In order to infer recurrent conformations across sampled sets, \OurTool{} agglomerate structure (multi)sets while keeping track of their condition of origin, and \Def{clusters} with respect to the \Def{base-pair distance}, the number of base pairs differing between two structures. A clustering algorithm then partitions the (multi)set of sampled structures into \emph{clusters}, (multi)sets of structures such that the accumulated sum of distances over clusters is minimized.

@@

Moreover, define the (pseudo-)\Def{Boltzmann condition probability} of a structure $S$, generated for a probing condition $d$ as part of a sampled  set $\mathcal{S}_d$, as
$$\mathbb{P}_d(S) = \frac{e^{-E_d(S)/RT}}{\mathcal{Z}^*_d} \text{, with } \mathcal{Z}^*_d := \sum_{S'\in \mathcal{S}_d} e^{-E_d(S')/RT}$$
where $E_d(S)$ is the pseudo free-energy, including the Turner free-energy, assigned to the structure $S$ within the probing condition $d$.
The \Def{stability} of a cluster $C$ denoted its accumulated pseudo-Boltzmann probability across conditions, computed as
$$\text{Stability}(C) = \sum_{d\in \mathcal{D}} \sum_{S\in C\cap \mathcal{S}_d}  \mathbb{P}_d(S).$$
A cluster is deemed \Def{significantly populated} if its stability exceeds a predefined threshold $\epsilon$. 
We set  $\epsilon =\RevA{|\mathcal{D}|}/3$ by default, such that at most three clusters are deemed significantly populated, and used as our primary candidates.
Finally, we consider two clusters to be \Def{highly similar} if their centroid structures differ by at most $\delta$ base pairs ($\delta=1$ by default), allowing the identification of clusters in the presence of minor variations.

@@

\paragraph{Filtering the promising conformer(s).} Finally, we \Def{elect the most promising cluster(s)}, and return their centroid(s). While the final number of clusters may potentially be large, only a handful of clusters are expected to represent structures that \RevA{are} both stable, and supported by a large number of conditions. The remaining clusters are indeed probably artifacts of the clustering method, but nevertheless useful to filter out \emph{noisy} structures.

\RevA{We postulate that a perfect cluster should have large stability, as defined above, and be representative of several conditions.} In the presence of a set of experimental conditions $\mathcal{D}$, we consider that a cluster $C$ supports a given condition $d$ when its probability within $d$ exceeds a given threshold $\tau$. The number of conditions supporting a cluster $C$ is defined as
$$\text{Support}(C) = \left|\left\{d\in \mathcal{D} \mid \left(\textstyle\sum_{S\in C\cap \mathcal{S}_d} \mathbb{P}_d(S)\right)\ge \tau \right\}\right|.$$
The value of $\tau$ is set to $1/(\NumClust+1)$ \RevA{with $\NumClust$ the final number of clusters},  ensuring at least one supporting cluster for each condition.

@@

\paragraph{Hajdin dataset.} 
A dataset was gathered by Hajdin\etal\cite{Hajdin2013} to validate the predictive capacities of probing data-driven predictions. It consists of $24$ RNA sequences with known secondary structures for which a single chemical probing reactivity profile (1M7-\SH) was made available.
This dataset includes sequences originating from a variety of organisms, and spans lengths ranging from $34$ nts to $530$ nts, with a focus on riboswitches and complex RNA architectures \RevA{(full list in Supp. Table S3)}.

@@

\paragraph{Didymium structural model and probing data (\didy{} dataset).} 
We considered the 188 nucleotides Lariat capping ribozyme from {\itshape Didymium iridis}, resolved 3.85 \AA{} resolution using X-ray cristallography (PDB: 4P8Z)~\citep{Meyer2014}.  We annotated the secondary structure elements using the \Software{DSSR} software from the 3DNA suite~\cite{Lu2015}. \RevA{Non-canonical base pairs were removed, and a non-pseudoknotted secondary structure was extracted as the maximum subset of non-crossing base pairs~\citep{Smit2008}.}

@@

\RevA{To systematically assess the potential of multiple sources of probing, we considered a difficult example, the \textit{Didymium iridis} Lariat Capping ribozyme (\didy{}). The native structure of \didy{}, shown in Figure~\ref{fig:didy} is highly complex, and features two pseudoknots which cannot be explicitly modeled by most computational methods, making \didy{} a challenging target for secondary structure prediction.}

\didy{} was probed with different SHAPE reagents: 1M7 (1-methyl-7 nitrosatoic anhydre), NMIA (N-Methyl Isatoic Anhydre), BzCN (Benzoyl cyanide) and NAI (2-methylnicotinic acid imidazolide) in presence and absence of Mg$^\text{2+}$. \RevA{\didy{} was also probed with DMS (DiMethylSulfate) and CMCT, in presence of Mg$^\text{2+}$, resulting in a total of 16 \Def{probing conditions}, a term we use in the following to denote a combination of probing technology, reagent and presence/absence of Mg$^\text{2+}$ (+ sequencing technology). For each probing condition, three experiments were performed in presence/absence of the reagent, and in a denatured context, following classic SHAPE protocols~\cite{Wilkinson2006,Smola2015}.}

@@

The Matthews Correlation Coefficient (MCC) is a classic metric for assessing the quality of a predicted structure $S$, identified by a set of base pairing positions, compared to an accepted reference (native) structure $\Ref$. It represents a compromise \RevA{between the main metrics derived from the confusion matrix}, and is defined as 
$$\text{MCC}(S\mid \Ref)=\frac{\text{TP} \times \text{TN} -\text{FP} \times \text{FN}}{\sqrt[]{(\text{TP}+\text{FP})(\text{TP}+\text{FN})(\text{TN}+\text{FP})(\text{TN}+\text{FN})}},$$
where TP, FP, TN, FN represent the correctly/erroneously predicted and correctly/erroneously omitted base pairs in $S$ with respect to $\Ref$.
\RevA{MCC is a stringent metric, taking values between -100\% and 100\%, 0\% being the expected MCC of a "coin tossing" random predictor.}

@@

\RevA{
\subsection*{Assessment of statistical significance}
Following Xu\etal\cite{Xu2012}, we assessed the statistical significance of observed changes in predictive performances using a two-tailed Student $t$-test (unequal variance) with usual significance threshold $\alpha=5\%$. When comparing performances within a given dataset, we used the paired version of the test.

@@

\subsection*{Considering multiple probing conditions \RevA{appears to} improve the quality of predictions}\label{sec:cordero}

@@

	\caption{Dot-bracket representations of the native structure and \OurTool{} predictions for the glycine riboswitch sequence of the Cordero\etal\cite{Cordero2012} data set, supplemented with an increasing collection of probing conditions. \RevA{Please note that, throughout the manuscript, by "conditions" we mean a combination of the nature of the probe, the probing technology, the ions present, the temperature, or any other "environmental" variation} Positions with green and red backgrounds indicate correctly and incorrectly predicted base pairs.}\label{fig:glycine_example}

@@

	\caption{\RevA{Predictive performances (GM) of \OurTool{} on the Cordero~\cite{Cordero2012a} dataset consisting of reactivities associated with the CMCT, NMIA and DMS probing for 6 reference RNAs. (Pseudo) MFE and MEA structures, predicted using \Software{RNAfold}~\cite{Lorenz2011}, are included as references in the case of a single condition, and in the absence of probing.}}
	\label{cordero1}
\end{figure}

@@

Remarkably, the combination of the three conditions leads to the best overall predictions, averaging $78.5\%$ GM.
This \RevA{seemingly} improves by $8.5\%$, $2.7\%$ and $1.4\%$ over the average predictions achieved using CMCT, DMS and NMIA respectively.
%CMCT/DMS/NMIA :0.6988333333/0.7588333333/0.771
Table~\ref{fig:glycine_example} illustrates the incremental refinement of \OurTool predictions for a glycine riboswitch upon increasing the number of considered probing conditions.
\RevA{However, these observations merely suggest an improvement, as statistical significance cannot be established on such a small dataset.}
Still, they suggests some level of complementarity between probing conditions, as already suggested by recent analyses~\citep{Brunel2000,Steen2012,Rice2014,Yu2018}  and supported by further analyses in this paper.

@@

\subsection*{\OurTool outperforms state-of-the-art predictive methods when informed by a single profile}
%Mono-probing Mathews (Hadjim)
%* On est meilleur sur la version courante de RSample
%* Multi conformation
We assessed the predictive capacities of \OurTool in a classic setting where a single probing condition is available, and compared its performances against \Software{RSample} software recently introduced by Spacic\etal\cite{Spasic2017}, \Software{RSample} relies on a sampling/clustering method developed independently from the current work. It was shown to perform \RevA{comparably as} a comprehensive collection of state-of-the-art methods in probing-guided structure prediction, including \Software{RME}~\citep{Wu2015}, \Software{RNAprob}~\citep{Deng2016}, \Software{RNAprobing}~\citep{Washietl2012}, \Software{RNAsc}~\citep{Zarringhalam2012} and the \Software{fold} utility from the \Software{RNAStructure} suite~\citep{Reuter2010}.
We considered the Hajdin dataset~\cite{Hajdin2013}, which consists of 24 RNAs of lengths ranging from 47 to 500 nts, all believed to fold into a unique documented conformation.

\begin{figure}
	\includegraphics[width=\linewidth]{graphs/RsampleVsIPANEMAP/Accuracy}
	\caption{\RevA{Comparison of predictive performances (GM) achieved by  \OurTool{} and \Software{Rsample} over the Hajdin\etal\cite{Hajdin2013} data set, consisting of 24 RNAs which associated SHAPE reactivities. The length of each individual RNA is indicated within square brackets.} \label{fig:Vsrsample}}
\end{figure}

\RevA{We used the default number of sampled structures for \Software{RSample}, namely 10~000 to compute correction factors and 1~000 for the clustering phase, and \OurTool (1~000 structures).  For all RNAs, a single structure was returned by both software. Note that the Pareto-optimality always implies a single returned cluster/structure for \OurTool{} in a mono probing setting.}
We computed and report in Figure~\ref{fig:Vsrsample} the individual Geometric Means (GM) of predicted structures. \RevA{Sensitivity and PPV can be found in Supp. Table S4, and Supp. Table S5 reports the stability of predictions over 10 independent runs of \OurTool.}

%\paragraph{Results description}
A first observation is that \OurTool generally outperforms \Software{RSample}, averaging $80\%$ GM as opposed to $68\%$ for its competitor. \RevA{The superiority on \OurTool on this benchmark appears statistically significant according to a two-tailed paired $t$-test (P-value$=6\times10^{-3}$) with observed data.}
Out of the 24 considered RNAs, \OurTool outperforms its competitor for 16 out of the 24 RNAs, achieves similar performances for 2 of them, and is dominated for the remaining 6 cases. Whenever dominant, \OurTool achieves an average GM score of $82\%$, dominating by $32.8\%$ its competitor, while being dominated by $3.5\%$ over the remaining examples (including ties). 
%Difference<-c(IPANEMAP-Rsample)
%> sum(Difference[1:18])
%[1] 3.17
%> sum(Difference[19:24])
%[1] -0.382
Remarkably, \OurTool predicts secondary structures much more consistently than its competitor. 
It ensures GM that are greater than 50\% in all cases, and even exceeds 60\% over all except for the yeast tRNA, encoding the Aspartic acid and HIV1 5' pseudoknot domain. The standard deviation of GM for \OurTool is $12.4\%$, compared to the larger value of $21\%$ for \Software{RSample} predictions.
